Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-05 paper/paper_draft.qmd
Updates to limitations section

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values that are tracked over
timesteps through growth. Additionally, we do not model the direct influence of light
intensity on coral growth or the negative influence of depth on coral growth rates due to
reduced available light levels [@baker1975; @huston1985; @ruiz-diaz2022]. If included in
this investigation it is possible that reef recovery ability, via regrowth, would be reduced
at greater depths, reducing the positive effect of depth refugia for temporal resilience.
Additionally, reduced growth rates at increased depths may reduce the carbonate budget
maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of
quality future projected data we do not consider other important impacts such as waves,
cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and
bleaching [@massel1993; @deaker2022].

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future
thermal stress projections are likely to influence the results obtained in this
investigation. While the resolution of connectivity data used in this investigation improves
upon data used in comparable studies [@bozec2025], the time-mean implementation used may not
effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity
implementation methods that better capture the ecological-timescale trends in reef
connections rather than the high daily variability, would improve confidence in the
connectivity findings of this investigation [@vogt-vincent2023]. Additionally greater year -
sample sizes for baseline connectivity data as future computation improvements are made,
along with the derivation of connectivity from an ensemble of biophysical models, will
improve confidence in connectivity data for ecological modelling and management applications
[@choukroun2025]. The DHW data used in this investigation have not been bias corrected for
the Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022]
developed using statistical downscaling. Additionally, Lachs et al. [@lachs2021] suggest the
utility of DHW levels for prediction of coral bleaching can be improved by reducing the
threshold for HotSpot Anomaly values from MMM +1 to MMM and reducing the accumulation window
from 12 weeks to 8 weeks. We investigate a range of future climate responses, thermal
projections and ecological parameterisations of ADRIA - CoralBlox to account for many
possible futures, and our results suggest the factors that influence reef resilience are
consistent across these scenarios.

Additionally, it is currently not possible to obtain highly accurate and precise estimates
of reef depth at the GBR-scale. Data adopted for the given assessment are said to be
accurate up to 15m from Mean Sea Level [@GBRMPA2020; @li2021], however there are many
sources of uncertainty in the depth estimates used for this study. Satellite-derived
estimates are subject to different bottom types (rock, sand, seagrass, etc) and their effect
on reflectance, the water penetration depth of satellite sensors, influence of local
conditions (e.g., turbidity and wave activity), among many other factors. The median depths
for each indicative reef area were extracted from bathymetry data using zonal statistics and
depth values beyond the stated 15m accuracy threshold [@GBRMPA2020] were included. The
resulting depth estimates are therefore highly uncertain and are only adopted as a useful
indication for the purpose of the presented analysis. They are not intended to be an
accurate and precise report of reef depths across the GBR.

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data of less than 5% of its indicative area were
excluded from analysis to ensure analyses were not conducted on reefs with no data coverage
(see @sfig-depth-cutoff). The existing GBRMPA reef outlines were adopted to align with other
ongoing modelling and analyses conducted as part of the Reef Restoration and Adaptation
Program (<https://gbrrestoration.org>; [@thereef]).

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model supports the robustness of our findings for understanding broad-scale
coral cover responses to climatic drivers. The aim here, however, was to assess the relative
influence of reef characteristics on reef resilience across a range of future conditions and
parameterisations, rather than provide spatially explicit and precise predictions of reef
outcomes. Using this approach, we suggest that reef depth is an important factor for reef
resilience under thermal stress.

Improvements in remote sensing approaches for broad scale bathymetry measurement and
understanding of how depth impacts experienced thermal stress on reefs will further enhance
investigations of depth as a critical factor for reef resilience. Additional analyses could
build on the findings of this study by assessing the impacts of prioritising reef depth in
reef intervention decision making. Future analyses could incorporate other environmental and
ecological stressors such as wave activity, ocean acidification, and Crown-of-Thorn Starfish
[@lin2023; @mcclanahan2005; @grimaldi2023; @grimaldi2022; @liu2024; @eyre2018].

>>>>

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. Coral colonies are modelled as diameter values that increase over time
through growth. Additionally, the influence of light intensity on coral growth or the
negative influence of depth on coral growth rates due to reduced available light levels are
not modelled in this investigation [@baker1975; @huston1985; @ruiz-diaz2022]. If included in
this investigation, it is possible that reef recovery ability, via regrowth, would be
reduced at greater depths, reducing the positive effect of depth refugia for temporal
resilience. Additionally, reduced growth rates at increased depths may reduce carbonate
budget maintenance due to lower rates of CaCO~3~ accretion. Due to the availability of
quality future projected data this investigation does not consider other important impacts
such as waves, cyclones, ocean currents and Crown of Thorns Starfish (CoTS) on coral colony
survival and bleaching [@massel1993; @deaker2022].

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future
thermal stress projections likely influence the results obtained in this investigation.
While the resolution of connectivity data used in this investigation improves upon data used
in comparable studies [@bozec2025], the time-mean implementation used may not effectively
capture long term patterns [@vogt-vincent2023]. Using alternative connectivity
implementation methods that better capture the ecological-timescale trends in reef
connections rather than the high daily variability, would improve confidence in the
connectivity findings of this investigation [@vogt-vincent2023]. Increased samples to inform
baseline connectivity data, along with the derivation of connectivity from an ensemble of
biophysical models, would also improve confidence in connectivity data for ecological
modelling and management applications [@choukroun2025].

DHW data used in this investigation have not been bias corrected for the Great Barrier Reef
region and are part of a wider Australian dataset [@dixon2022] developed using statistical
downscaling. Lachs et al. [-@lachs2021] suggest the utility of DHW levels for prediction of
coral bleaching can be improved by reducing the threshold for HotSpot Anomaly values from
MMM +1 to MMM and reducing the accumulation window from 12 weeks to 8 weeks. The presented
analyses assessed a range of future climate responses, thermal projections and ecological
parameterisations of ADRIA-CoralBlox to account for many possible futures and suggests the
factors that influence reef resilience are consistent across these scenarios.

Highly accurate and precise estimates of reef depth across the GBR are currently not
possible to obtain with (publicly) available data. Data adopted for the given assessment are
said to be accurate up to 15m from Mean Sea Level [@GBRMPA2020; @li2021], however there are
many sources of uncertainty in the depth estimates used for this study. Satellite-derived
estimates are subject to different bottom types (rock, sand, seagrass, etc) and their effect
on reflectance, the water penetration depth of satellite sensors, influence of local
conditions (e.g., turbidity and wave activity), among many other factors. The median depths
for each indicative reef area were extracted from bathymetry data using zonal statistics and
depth values beyond the stated 15m accuracy threshold [@GBRMPA2020] were included. The
resulting depth estimates are therefore highly uncertain and are only adopted as a useful
indication for the purpose of the presented analysis. They are not intended to be accurate
and precise reef depths across the GBR.

In addition to the usual challenges of deriving data from satellite imagery, it is
acknowledged that there is high potential for misalignment between the bathymetry data and
GBRMPA reef outlines. Hammerton and Lawrey [-@hammerton2025], for example, identified 120
false positives, where features identified as reefs were sand banks or rocky reefs (rather
than coral reefs). Reefs with bathymetry data covering less than 5% of their indicative area
were excluded from analyses (see @sfig-depth-cutoff). Existing GBRMPA reef outlines were
adopted to align with other ongoing modelling and analyses conducted as part of the Reef
Restoration and Adaptation Program (<https://gbrrestoration.org>; [@thereef]).

The modelling conducted with ADRIA-CoralBlox itself is subject to many uncertainties. For
one, a formal mechanistic and spatial validation of the model represents an area for future
refinement and is the focus of a future publication. The consistency of projected patterns
with known coral-climate relationships and reef characteristics provides confidence in the
model's ability to meaningfully capture ecological signals at the GBR-scale. The mechanistic
basis of the model supports the robustness of these findings for understanding broad-scale
coral cover responses to climatic drivers. The aim here, however, was to assess the relative
influence of reef characteristics on reef resilience across a range of future conditions and
parameterisations, rather than provide spatially explicit and precise predictions of reef
outcomes. Using this approach, results suggest reef depth is an important factor for reef
resilience under thermal stress.

Improvements in remote sensing approaches for broad scale bathymetry measurement and
understanding of how depth impacts experienced thermal stress on reefs will further enhance
investigations of depth as a critical factor for reef resilience. Additional analyses could
build on the findings of this study by assessing the impacts of prioritising reef depth for
reef intervention decision outcomes. Future analyses could incorporate other environmental
and ecological stressors such as wave activity, ocean acidification, and Crown-of-Thorn
Starfish [@lin2023; @mcclanahan2005; @grimaldi2023; @grimaldi2022; @liu2024; @eyre2018].
